Reasonably Programmable Literal Notation

### 2018-06-22 icfp18/syntax-icfp18.tex
revised 2.2.2

\subsubsection{\li{RegexLexer}}
\label{sec:regex-lexer}
% A fragment of the signature of the \li{Relit_helper} library, which provides various necessary definitions and utilities to TLM providers, appears in Fig.~\ref{fig:relit-util}. 
In most respects, the definition of the lexer in Fig.~\ref{fig:regex-lexer} is conventional. It reads various token patterns on Lines 8-32, emitting the corresponding tokens specified by the \lstinline[morekeywords={token}]{%token} declarations in Fig.~\ref{fig:regex-parser}, which have been brought into scope by Line 2 of Fig.~\ref{fig:regex-lexer}. Lines 22-23 combine sequences of non-special characters (e.g. alphanumeric characters) into a single string token. Lines 24-25 implement backslash-prefixed escape sequences for characters that have special meaning in this notation for regexes, emitting a string token in each instance.\footnote{A better approach would be to define a second lexing rule that greedily combines sequences consisting of non-special characters and escape characters into a single \li{STR} token, but for the sake of exposition, we stick to this simpler approach.}

Lines 24-31 of the lexer, which recognize the notation for regex splicing \li{$(}$e$\li{)}, and string splicing, \li{$$(}$e$\li{)}, are more unusual. Recall from Sec.~\ref{sec:context-dependence} that the parser must represent spliced terms abstractly, by their location within the literal body. This implies that the TLM does not need to parse spliced expressions itself, as long as it can independently decide where the spliced expressions start and end. A paired start and end position is a segment,  represented by a value of the type \li{Segment.t} defined in the \li{Relit} helper library as specified on Line 2 of Fig.~\ref{fig:relit-util}. Lines 26-28 and Lines 29-31 of the lexer produce a segment by calling a helper function, \li{Relit.Segment.read_to}, that internally invokes the Reason lexer on the provided lexing buffer until it sees an instance of the provided token outside of a Reason comment, string literal or generalized literal. If the provided token is a right delimiter also used by Reason, like \li{")"} on Line 4 of Fig.~\ref{fig:regex-lexer}, then \li{read_to} looks for the first unmatched instance. For example, if the remaining lexing buffer contains {\small \texttt{f("))"))A|G}} then \li{read_to} will consume up until just before the final instance of \li{)} and emit the appropriate segment. Similar helper functions, not shown, are available for reading out single identifiers, simple paths like \li{X1.X2.xy} and a few other unambiguously delimited fragments of the Reason syntax.

This approach has two major benefits. First, it supports splicing even within literal notations that have a very different lexical structure from Reason itself, as in this example. Second, it allows the parser to avoiding needing to link to Reason's expression grammar (though this is supported by Menhir), which substantially simplifies reasoning about ambiguities. Spliced expressions arrive as single, opaque tokens as specified on Lines 6-7 of Fig.~\ref{fig:regex-parser}. The system will recursively process spliced expressions after the parser has finished generating the proto-expansion.

@@

In order to repurpose this existing parse tree representation for proto-expansions, we must provide some way to unambiguously represent splice references, notated earlier in the paper as \li{spliced<m..n : ty>} where \li{m..n} specifies a segment and \li{ty} is the segment's expected type. The \li{ProtoExpr} module solves this problem by providing a function \li{spSURLEURLliced} that takes a segment, of type \li{Segment.t}, and a type representation of type \li{Parsetree.core_type}, which also comes from OCaml's compiler library, and produces a corresponding value of type \li{ProtoExpr.t} that uniquely represents the corresponding splice reference (see Sec.~\ref{sec:implementation} for more on the internal representation).\todo{say something about splice references in sec 6}{}

The datatypes in OCaml's \li{Parsetree} module are necessarily intricate, given the sophistication of the OCaml system, so constructing expression encodings manually is generally too unwieldy. Addressing this class of problem is, of course, the motivation for this very paper, so it is only natural that we define TLMs  for working with OCaml parse trees using familiar surface syntax, extended with splice forms. In the literature, literal notation for the host language's own parse trees is called \emph{quasiquotation}, and splicing is referred to as \emph{unquotation} or \emph{antiquotation} \cite{Bawd99a,shabalin2013quasiquotes,mainland2007s}. The TLM \li{$re_expr}  specified on Line 10 of Fig.~\ref{fig:relit-util} provides quasiquotation for OCaml parse trees using Reason's surface syntax. The accompanying technical report describes its implementation. For clarity, the \li{$proto_expr} TLM is defined as a synonym for \li{$re_expr} on Line~12 of Fig.~\ref{fig:relit-util} (the two TLMs may diverge in the future). We apply this TLM many times in the semantic actions in Fig.~\ref{fig:regex-parser}. Notice, however, that the generalized literal forms in Fig.~\ref{fig:regex-parser} are not each prefixed by \li{$proto_expr}. Instead, we use the \li{open notation} directive on Line 2, which implicitly applies \li{$proto_expr} to all unadorned generalized literal forms following the usual lexical scoping rules for \li{open}.

@@

This issue does \textbf{not} come up when using TLMs to define quotation literals for languages other than Reason itself (an example is given in Sec.~\ref{sec:sptsms}), so if this design proves too limiting, it may be reasonable to build ``inside-out'' quasiquotation primitively into Reason as a special case. In fact, this is already possible using MetaOCaml, discussed below. %(Nested TLM application via splicing is, of course, useful and fully supported, e.g. in Fig.~\ref{fig:nesting-example}).
 % Typed quotations with an explicit representation of contexts would need to be built in to the language  necessary to even think about expanding a

 \todo{integrate me}{}Another benefit of this explicit approach is that renamings can be performed without needing to reason about or modify parse tree encodings in parser definitions. For example, if we decided to rename the module \li{Regex} to \li{Regexp}, this renaming would need only to be performed on the module paths in the TLM definition, as long as the name of the internal module variable in the dependencies list remains \li{Regex}. 
%(The TLM provider may wish, for the sake of aesthetics, to propagate the change into the parser definition as well, but this is not semantically necessary). 
This allows us to encode expansions using ordinary OCaml values (in fact, using the OCaml compiler's own parse tree representation), rather than requiring the introduction of primitive quasiquotation into OCaml (see Sec. \ref{sec:provider-perspective}).

>>>>

\subsubsection{\li{RegexLexer}}
\label{sec:regex-lexer}
% A fragment of the signature of the \li{Relit_helper} library, which provides various necessary definitions and utilities to TLM providers, appears in Fig.~\ref{fig:relit-util}. 
In most respects, the definition of the lexer in Fig.~\ref{fig:regex-lexer} is conventional. It reads various lexical patterns on Lines 8-32, emitting the corresponding tokens specified by the \lstinline[morekeywords={token}]{%token} declarations in Fig.~\ref{fig:regex-parser}, which have been brought into scope by Line 2 of Fig.~\ref{fig:regex-lexer}. Lines 22-23 combine sequences of non-special characters (e.g. alphanumeric characters) into a single string token. Lines 24-25 implement backslash-prefixed escape sequences for characters that have special meaning in this notation for regexes, emitting a string token in each instance.\footnote{A better approach would be to define a second lexing rule that greedily combines sequences consisting of non-special characters and escape characters into a single \li{STR} token, but for the sake of exposition, we stick to this simpler approach.}

Lines 24-31 of the lexer, which recognize the notation for regex splicing \li{$(}$e$\li{)}, and string splicing, \li{$$(}$e$\li{)}, are more unusual. Recall from Sec.~\ref{sec:context-dependence} that the parser must represent spliced terms abstractly, by their location within the literal body. This implies that the TLM does not need to parse the spliced expression itself, as long as it can independently decide where the spliced expression starts and ends. A paired start and end position is a segment,  represented by a value of the type \li{Segment.t} defined in the \li{Relit} helper library as specified on Line 2 of Fig.~\ref{fig:relit-util}. Lines 26-28 and Lines 29-31 of the lexer produce a segment by calling a helper function, \li{Relit.Segment.read_to}, that internally invokes the Reason lexer on the provided lexing buffer until it sees an instance of the provided token outside of a Reason comment, string literal or generalized literal. If the provided token is a right delimiter also used by Reason, like \li{")"} on Line 4 of Fig.~\ref{fig:regex-lexer}, then \li{read_to} looks for the first unmatched instance. For example, if the remaining lexing buffer contains {\small \texttt{f("))"))A|G}} then \li{read_to} will consume up until just before the final instance of \li{)} and emit the appropriate segment. Similar helper functions, not shown, are available for reading out single identifiers, simple paths like \li{X1.X2.xy} and a few other unambiguously delimited fragments of the Reason syntax.

This approach has two major benefits. First, it supports splicing even within literal notations that have a very different lexical structure from Reason itself, as in this example. Second, it allows the parser to avoiding needing to link to Reason's expression grammar (though this is supported by Menhir), which substantially simplifies reasoning about ambiguities. Spliced expressions arrive into the parser as single, opaque tokens as specified on Lines 6-7 of Fig.~\ref{fig:regex-parser}. The system will recursively expand spliced expressions after the parser has finished generating the proto-expansion.

@@

In order to repurpose this existing parse tree representation for proto-expansions, we must provide some way to unambiguously represent splice references, notated earlier in the paper as \li{spliced<m..n : ty>} where \li{m..n} specifies a segment and \li{ty} is the segment's expected type. The \li{ProtoExpr} module solves this problem by providing a function \li{spSURLEURLliced} that takes a segment, of type \li{Segment.t}, and a type representation, of type \li{Parsetree.core_type}, which also comes from OCaml's compiler library, and produces a corresponding value of type \li{ProtoExpr.t} that uniquely represents the corresponding splice reference (see Sec.~\ref{sec:implementation} for more on the internal representation).\todo{say something about splice references in sec 6}{}

Note that it is only because we explicitly name dependencies in TLM definitions that we can get away with using a simple, unprivileged representation of parse trees. In particular, 
% Another benefit of this explicit approach is that 
variable renamings can be performed without needing to reason about or modify parse tree encodings in parser definitions (where variables are represented using strings, and are thus not part of the binding structure). For example, if we decided to rename the module \li{Regex} to \li{Regexp}, this renaming would need only to be performed on the module paths in the definition of \li{$rx}, as long as the name of the internal module variable in the dependencies list remains \li{Regex}. 
%(The TLM provider may wish, for the sake of aesthetics, to propagate the change into the parser definition as well, but this is not semantically necessary). 
% This allows us to encode expansions using ordinary OCaml values (in fact, using the OCaml compiler's own parse tree representation), rather than requiring the introduction of primitive quasiquotation into OCaml (see Sec. \ref{sec:provider-perspective}).

The datatypes in OCaml's \li{Parsetree} module are necessarily intricate, given the sophistication of the OCaml system, so constructing expression encodings manually is generally too unwieldy. Addressing this class of problem is, of course, the motivation for this very paper, so it is only natural that we define TLMs  for working with OCaml parse trees using familiar surface syntax, extended with splice forms. In the literature, literal notation for the host language's own parse trees is called \emph{quasiquotation}, and splicing is referred to as \emph{unquotation} or \emph{antiquotation} \cite{Bawd99a,shabalin2013quasiquotes,mainland2007s}. The TLM \li{$re_expr}  specified on Line 10 of Fig.~\ref{fig:relit-util} provides quasiquotation for OCaml parse trees using Reason's surface syntax. The accompanying technical report describes its implementation. For clarity, the \li{$proto_expr} TLM is defined as a synonym for \li{$re_expr} on Line~12 of Fig.~\ref{fig:relit-util} (the two TLMs may diverge in the future). We apply this TLM many times in the semantic actions in Fig.~\ref{fig:regex-parser}. Notice, however, that the generalized literal forms in Fig.~\ref{fig:regex-parser} are not each prefixed by \li{$proto_expr}. Instead, we use the \li{open notation} directive on Line 2, which implicitly applies \li{$proto_expr} to all unadorned generalized literal forms in its scope.

@@

This issue does \textbf{not} come up when using TLMs to define quotation literals for languages other than Reason itself (an example is given in Sec.~\ref{sec:sptsms}), so if this design proves too limiting, it may be reasonable to build ``inside-out'' quasiquotation primitively into Reason. In fact, this is already possible using a preprocessor, \li{ppx_metaquot}, from the \li{ppx_tools} library \cite{ppx-tools}, or with MetaOCaml and related systems, discussed below. %(Nested TLM application via splicing is, of course, useful and fully supported, e.g. in Fig.~\ref{fig:nesting-example}).
 % Typed quotations with an explicit representation of contexts would need to be built in to the language  necessary to even think about expanding a
